---
title: "Anisotropic models"
author: "David Bolin and Alexandre B. Simas"
date: "Created: 2024-10-23. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Anisotropic models}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---
    
```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(rSPDE)
library(INLA)    
set.seed(123)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
    sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction

For domains $D\subset \mathbb{R}^2$, the `rSPDE` package implements the anisotropic Matérn model
$$
    (I - \nabla\cdot (H\nabla))^{(\nu + 1)/2} u = c\sigma W, \quad \text{on }  D
$$
Where $H$ is a $2\times 2$ positive definite matrix, $\sigma, \nu >0$ and $c$ is a constant chosen such that $u$ would have 
the covariance function 
$$
r(h) = \frac{\sigma^2}{2^{\nu-1}\Gamma(\nu)} (\sqrt{h^T H^{-1} h})^{\nu} K_{\nu}(\sqrt{h^T H^{-1} h}),
$$
if the domain was $D = \mathbb{R}^2$, i.e., a stationary and anisotropic Matérn covariance function. The matrix $H$ is defined as 
$$
H = \begin{bmatrix}
h_x^2 & h_xh_y h_{xy}\\
h_xh_y h_{xy} & h_y^2
\end{bmatrix}
$$
with $h_x,h_y>0$ and $h_{xy} \in (-1,1)$. 

## Implementation details 
Let us begin by loading some packages needed for making the plots
```{r library_loading, message=FALSE}
library(ggplot2)
library(gridExtra)
library(viridis)
```
We start by creating a region of interest and a spatial mesh using the `fmesher` package:
```{r}
library(fmesher)
n_loc <- 2000
loc_2d_mesh <- matrix(runif(n_loc * 2), n_loc, 2)
mesh_2d <- fm_mesh_2d(
    loc = loc_2d_mesh,
    cutoff = 0.01,
    max.edge = c(0.1, 0.5)
)
plot(mesh_2d, main = "")
```

We now specify the model using the `matern2d.operators` function. 
```{r}
sigma <- 1
nu = 0.5
hx <- 0.08
hy <- 0.08
hxy <- 0.5
op <- matern2d.operators(hx = hx, hy = hy, hxy = hxy, nu = nu, 
                         sigma = sigma, mesh = mesh_2d)
```

The `matern2d.operators` object has `cov_function_mesh` method which can be used 
evaluate the covariance function on the mesh. For example
```{r}
r <- op$cov_function_mesh(matrix(c(0.5,0.5),1,2))
proj <- fm_evaluator(mesh_2d, dims = c(100, 100), xlim = c(0,1), ylim = c(0,1))
r.mesh <- fm_evaluate(proj, field = as.vector(r))
cov.df <- data.frame(x1 = proj$lattice$loc[,1],
                     x2 = proj$lattice$loc[,2], 
                     cov = c(r.mesh))
ggplot(cov.df, aes(x = x1, y = x2, fill = cov)) + geom_raster() + xlim(0,1) + ylim(0,1) + scale_fill_viridis()
```


We can simulate from the field using the `simulate method`:
```{r}
u <- simulate(op)
proj <- fm_evaluator(mesh_2d, dims = c(100, 100), xlim = c(0,1), ylim = c(0,1))
u.mesh <- fm_evaluate(proj, field = as.vector(u))
cov.df <- data.frame(x1 = proj$lattice$loc[,1],
                     x2 = proj$lattice$loc[,2], 
                     u = c(u.mesh))
ggplot(cov.df, aes(x = x1, y = x2, fill = u)) + geom_raster() + xlim(0,1) + ylim(0,1) + scale_fill_viridis()
```

Let us now simulate some data based on this simulated field.

```{r}
n.obs <- 2000
obs.loc <- cbind(runif(n.obs),runif(n.obs))
A <- fm_basis(mesh_2d,obs.loc)
sigma.e <- 0.1
Y <- as.vector(A%*%u + sigma.e*rnorm(n.obs))
```

We can compute kriging predictions of the process $u$ based on these observations. 
To specify which locations that should be predicted, the argument `Aprd` is used. 
This argument should be an observation matrix that links the mesh locations to the prediction locations.  
```{r, fig.show='hold',fig.align = "center"}
A <- op$make_A(obs.loc)
Aprd <- op$make_A(proj$lattice$loc)
u.krig <- predict(op, A = A, Aprd = Aprd, Y = Y, sigma.e = sigma.e)
```
The process simulation,  and the kriging prediction are shown in the following figure.

```{r, fig.show='hold', fig.align = "center",echo=TRUE}
data.df <- data.frame(x = obs.loc[,1], y = obs.loc[,2], field = Y, type = "Data")
krig.df <- data.frame(x = proj$lattice$loc[,1], y = proj$lattice$loc[,2],
                      field = as.vector(u.krig$mean), type = "Prediction")
df_plot <- rbind(data.df, krig.df)
ggplot(df_plot) + aes(x = x, y = y, fill = field) +
        facet_wrap(~type) + geom_raster(data = krig.df) +
        geom_point(data = data.df, aes(colour = field),
        show.legend = FALSE) + 
        scale_fill_viridis() + scale_colour_viridis()

```
Let us finally use `rspde_lme` to estimate the parameters based on the data.

```{r}
df <- data.frame(Y = as.matrix(Y), x = obs.loc[,1], y = obs.loc[,2])
res <- rspde_lme(Y ~ 1, loc = c("x","y"), data = df, model = op, parallel = TRUE)
```

In the call, `y~1` indicates that we also want to estimate a mean value of the model, and the arguments `loc` provides the names of the spatial and temporal coordinates in the data frame. 
Let us see a summary of the fitted model:

```{r}
summary(res)
```

Let us compare the estimated results with the true values:
```{r}
results <- data.frame(sigma = c(sigma, res$coeff$random_effects[2]),
                      hx = c(hx, res$coeff$random_effects[3]), 
                      hy = c(hy, res$coeff$random_effects[4]),
                      hxy = c(hxy, res$coeff$random_effects[5]),
                      nu = c(nu, res$coeff$random_effects[1]),
                      sigma.e = c(sigma.e, res$coeff$measurement_error),
                      intercept = c(0, res$coeff$fixed_effects),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

Finally, we can also do prediction based on the fitted model as 
```{r}
pred.data <- data.frame(x = proj$lattice$loc[,1], y = proj$lattice$loc[,2])
pred <- predict(res, newdata = pred.data, loc = c("x","y"))

data.df <- data.frame(x = obs.loc[,1], y = obs.loc[,2], field = Y, type = "Data")
krig.df <- data.frame(x = proj$lattice$loc[,1], y = proj$lattice$loc[,2],
                      field = as.vector(u.krig$mean), type = "Prediction")
df_plot <- rbind(data.df, krig.df)
ggplot(df_plot) + aes(x = x, y = y, fill = field) +
        facet_wrap(~type) + geom_raster(data = krig.df) +
        geom_point(data = data.df, aes(colour = field),
        show.legend = FALSE) + 
        scale_fill_viridis() + scale_colour_viridis()

```

## Inlabru Implementation

We will now fit the model using the `inlabru` package. 
First, we load the package, and create the anisotropic model object:

```{r}
library(inlabru)

model_aniso <- rspde.anistropic2d(mesh = mesh_2d)
```

Then we create the component and fit the model using the `inlabru` package. Observe that we can use the same data frame as in the `rspde_lme` example.
```{r}
cmp <- Y ~ Intercept(1) - 1 + field(cbind(x,y), model = model_aniso)
bru_fit_field <- bru(cmp, data = df, options = list(verbose = TRUE))
```

Let us now compare the estimated results (the means of the parameters) with the true values. Observe that we are using the helper function `transform_parameters_anisotropic()`
to change the scale of the estimated parameters back to the original scale.

```{r}
results <- data.frame(
  rbind(
    # True values
    c(hx = hx, hy = hy, hxy = hxy, sigma = sigma, nu = nu),
    # Estimated values transformed using transform_parameters_anisotropic
    transform_parameters_anisotropic(
      bru_fit_field$summary.hyperpar$mean[2:6],
      model_aniso[["nu_upper_bound"]]
    )
  ),
  # Add sigma.e values
  sigma.e = c(sigma.e, sqrt(1 / bru_fit_field$summary.hyperpar$mean[1])),
  # Add intercept values
  intercept = c(0, bru_fit_field$summary.fixed$mean),
  row.names = c("True", "Estimate")
)

print(results)
```