---
title: "Building the rSPDE package from source on Mac and Linux"
author: "David Bolin and Alexandre B. Simas"
date: "Created: 2023-01-17. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building the rSPDE package from source on Mac and Linux}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Starting with this release of `rSPDE`, the C/C++ `cgeneric` sources used by
`INLA` live on a single branch (`devel`) together with the R code, and
compilation is **opt-in**. The default install — including the CRAN release
and the regular GitHub install — is pure R and requires no compiler.

When you opt in, the `configure` script copies the sources from
`inst/src-optional/` into `src/` and then `R CMD INSTALL` builds
`inst/shared/rspde_cgeneric_models.so` (or `.dll` on Windows). At runtime
`rSPDE` detects this shared object and uses it directly; otherwise it falls
back to the shared object that ships with `INLA`. The legacy `devel-src` and
`stable-src` branches are deprecated.

## Asking for the compiled build

Pick one of the following — both have the same effect:

```{r, eval=FALSE}
# Via configure args
remotes::install_github(
  "davidbolin/rSPDE",
  ref = "devel",
  configure.args = "--enable-compiled"
)
```

```{r, eval=FALSE}
# Via an environment variable
withr::with_envvar(
  c(RSPDE_COMPILE = "1"),
  remotes::install_github("davidbolin/rSPDE", ref = "devel")
)
```

Without the flag (or `RSPDE_COMPILE=1`), `configure` writes a no-op `Makevars`
and the install stays pure R.

If you have already cloned the repository and want to build locally:

```{bash, eval=FALSE}
git clone https://github.com/davidbolin/rSPDE.git
cd rSPDE
RSPDE_COMPILE=1 R CMD INSTALL .
```

You can verify the compiled path is active from R with:

```{r, eval=FALSE}
getOption("rspde.compiled")  # TRUE when the local shared object is loaded
```

## Toolchain prerequisites

The `Makefile` shipped in `inst/src-optional/` is the same one that used to
live on `devel-src`. It expects:

- **Eigen** headers (the `cgeneric` C++ helpers `#include <Eigen/...>`),
- **`gcc-14` / `g++-14`** (Homebrew on macOS, distro packages on Linux),
- **BLAS / LAPACK** (linked via `-lblas -llapack`),
- on macOS, the Homebrew prefix exposed at the symlinked locations
  `/usr/local/brewlib` and `/usr/local/brewinclude`.

### Linux

Install Eigen:

```{bash, eval=FALSE}
sudo apt install libeigen3-dev          # Debian / Ubuntu
sudo pacman -S eigen                    # Arch / Manjaro
sudo yum install eigen3-devel           # RHEL / Fedora / CentOS
sudo zypper install eigen3-devel        # openSUSE
```

Install `gcc-14` / `g++-14` (or the most recent `gcc` available; the Makefile
defaults to `gcc-14`):

```{bash, eval=FALSE}
sudo apt install gcc-14 g++-14          # Debian / Ubuntu
sudo pacman -S gcc                      # Arch / Manjaro
sudo dnf install gcc gcc-c++            # Fedora
sudo zypper install gcc gcc-c++         # openSUSE
```

If your `gcc` is not at `/opt/homebrew/bin/gcc-14`, edit the `CC` / `CXX`
lines in `inst/src-optional/Makefile` (or in `src/Makefile` after running
`configure --enable-compiled`) before running `R CMD INSTALL`. The Makefile
auto-detects Linux and emits the correct architecture flags.

### macOS

Install [Homebrew](https://brew.sh/):

```{bash, eval=FALSE}
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
```

Install Eigen, `gcc@14`, and the Homebrew copies of BLAS/LAPACK that the
package links against:

```{bash, eval=FALSE}
brew install eigen
brew install gcc@14
```

If `eigen` is not picked up automatically, create the symlinks the Makefile
looks for:

```{bash, eval=FALSE}
sudo ln -s $(brew --prefix eigen)/include/eigen3 /usr/local/include/eigen3
sudo ln -s $(brew --prefix)/lib /usr/local/brewlib
sudo ln -s $(brew --prefix)/include /usr/local/brewinclude
```

The Makefile sets `CC = /opt/homebrew/bin/gcc-14` and
`CXX = /opt/homebrew/bin/g++-14`. On Apple Silicon (Homebrew under
`/opt/homebrew`) the path is correct out of the box. On Intel Macs (Homebrew
under `/usr/local`) either edit those two lines to point at your `gcc-14`
binary, or expose a symlink:

```{bash, eval=FALSE}
sudo ln -s $(brew --prefix gcc@14)/bin/gcc-14 /opt/homebrew/bin/gcc-14
sudo ln -s $(brew --prefix gcc@14)/bin/g++-14 /opt/homebrew/bin/g++-14
```

The Makefile also `codesign`s the produced `.so` ad-hoc, which is required by
the macOS dynamic linker on macOS 26+.

## Inspecting and adjusting the Makefile

The Makefile shipped in `inst/src-optional/` is reproduced (truncated) below.
After running `./configure --enable-compiled` you will find the same file in
`src/`, where any tweaks for an unusual environment should be made before
running `R CMD INSTALL .`:

```{bash, eval=FALSE}
toInclude = $(R_LIBRARY_DIR)/INLA/include/
PKG_CFLAGS = -I$(R_HOME)/include

CC  = /opt/homebrew/bin/gcc-14
CXX = /opt/homebrew/bin/g++-14

EIGEN_MAC     = /opt/homebrew/
EIGEN_MAC_VAR = /usr/local
EIGEN_LINUX   = /usr
MAC_LIB       = /usr/local/brewlib
MAC_INCLUDE   = /usr/local/brewinclude

# --- platform-specific flags follow; see the file for the full content ---
```

Most users only need to update `CC` / `CXX` (if `gcc-14` lives elsewhere) and
the Eigen / Homebrew paths. The first time you build, the Makefile will
fetch `cgeneric.h` from the upstream `r-inla` repository if it is not
already present in `${R_LIBRARY_DIR}/INLA/include/`.

## Windows

The Makefile is not maintained for Windows. Windows users should stick with
the default pure-R install (`remotes::install_github("davidbolin/rSPDE", ref = "devel")`),
which delegates the cgeneric implementation to the shared object that ships
with `INLA`.
