---
title: "Intrinsic models in the rSPDE package"
author: "David Bolin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intrinsic models in the rSPDE package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: xiong22
  title: "Covariance-based rational approximations of fractional SPDEs for computationally efficient Bayesian inference"
  author: 
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Xiong
    given: Zhen    
  container-title: Journal of Computational and Graphical Statistics
  type: article-journal
  issued:
    year: 2023   
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

set.seed(1)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction

In this vignette we provide a brief introduction to the intrinsic models implemented in the `rSPDE`
package. 

## Model specification and simulation 

The intrinsic models are defined as  
$$
(-\Delta)^{\beta/2}(\kappa^2-\Delta)^{\alpha/2}(\tau u) = \mathcal{W},
$$ 
where $\alpha + \beta > d/2$ and $d$ is the dimension of the spatial domain. These models are handled 
by performing two rational approximations, one for each fractional operator. 

To illustrate these models, we begin by defining a mesh over $[0,1]\times [0, 1]$:

```{r, message=FALSE, warning=FALSE, fig.align='center'}
library(fmesher)
bnd <- fm_segm(rbind(c(0, 0), c(2, 0), c(2, 2), c(0, 2)), is.bnd = TRUE)
mesh_2d <- fm_mesh_2d(
    boundary = bnd, 
    cutoff = 0.04,
    max.edge = c(0.05)
)
plot(mesh_2d, main = "")

```

We now use the `intrinsic.matern.operators()` function to construct the `rSPDE` representation
of the model. 

```{r, message=FALSE}
library(rSPDE)
kappa <- 10
tau <- 0.01
alpha <- 2
beta <- 1
op <- intrinsic.matern.operators(kappa = kappa, tau = tau, alpha = alpha, 
                                 beta = beta, mesh = mesh_2d)
```

To see that the `rSPDE` model is approximating the true model, we can compare the variogram
of the approximation with the true variogram (implemented in `variogram.intrinsic.spde()`) as follows.
```{r, message=FALSE}
Sigma <- op$A[,-1] %*% solve(op$Q[-1,-1], t(op$A[,-1]))
One <- rep(1, times = ncol(Sigma))
D <- diag(Sigma)
Gamma <- 0.5 * (One %*% t(D) + D %*% t(One) - 2 * Sigma)
point <- matrix(c(1,1),1,2)
Aobs <- spde.make.A(mesh = mesh_2d, loc = point)
vario <- variogram.intrinsic.spde(point, mesh_2d$loc[,1:2], kappa, 
                                  alpha = alpha, tau = tau,
                                  beta = beta, L = 2, d = 2)

d = sqrt((mesh_2d$loc[,1]-point[1])^2 +  (mesh_2d$loc[,2]-point[2])^2)
plot(d, Aobs%*%Gamma, xlim = c(0,0.5), ylim = c(0,0.2))
lines(sort(d),sort(vario),col=2, lwd = 2)
```

We can now use the `simulate` function to simulate a realization of the field $u$:

```{r}
u <- simulate(op,nsim = 1)

proj <- fm_evaluator(mesh_2d, dims = c(100, 100))
field <- fm_evaluate(proj, field = as.vector(u))
field.df <- data.frame(x1 = proj$lattice$loc[,1],
                       x2 = proj$lattice$loc[,2], 
                       y = as.vector(field))

library(ggplot2)
library(viridis)
ggplot(field.df, aes(x = x1, y = x2, fill = y)) +
    geom_raster() +
    scale_fill_viridis()

```

By default, the field is simulated with a zero-integral constraint. 


Let us now consider a simple Gaussian linear model where 
the spatial field $u(\mathbf{s})$ is observed at
$m$ locations, $\{\mathbf{s}_1 , \ldots , \mathbf{s}_m \}$ under Gaussian measurement noise.
For each $i = 1,\ldots,m,$ we have
$$
\begin{align} 
y_i &= u(\mathbf{s}_i)+\varepsilon_i\\
\end{align},
$$
where $\varepsilon_1,\ldots,\varepsilon_{m}$ are iid normally distributed
with mean 0 and standard deviation 0.1.

To generate a data set `y` from this model, we first draw some observation locations at random in the domain and then
use the `spde.make.A()` functions (that wraps the functions `fm_basis()`, `fm_block()` and `fm_row_kron()` of the `fmesher` package) to construct the observation matrix which can be used to evaluate the
simulated field $u$ at the observation locations. After this we simply add the measurment noise. 

```{r}
n_loc <- 2000
loc_2d_mesh <- matrix(runif(n_loc * 2), n_loc, 2)

A <- spde.make.A(
  mesh = mesh_2d,
  loc = loc_2d_mesh
)
sigma.e <- 0.1
y <- A %*% u + rnorm(n_loc) * sigma.e
```

The generated data can be seen in the following image.

```{r,fig.align = "center", echo=TRUE}
library(ggplot2)
library(viridis)
df <- data.frame(x1 = as.double(loc_2d_mesh[, 1]),
  x2 = as.double(loc_2d_mesh[, 2]), y = as.double(y))
ggplot(df, aes(x = x1, y = x2, col = y)) +
  geom_point() +
  scale_color_viridis()
```



## Fitting the model with `R-INLA` 

We will now fit the model using our `r inla_link()` implementation of the
rational SPDE approach. Further details on this implementation can be found in
 [R-INLA implementation of the rational SPDE approach](rspde_inla.html).
 
We begin by loading the `INLA` package and creating the $A$ matrix, the index, and the
`inla.stack` object. For now, the model can only be estimated with $\beta = 1$ and
$\alpha = 1$ or $\alpha = 2$. For these non-fractional models, we can use the standard
INLA functions to make the required elements. 

```{r}
library(INLA)
mesh.index <- inla.spde.make.index(name = "field", n.spde = mesh_2d$n)

st.dat <- inla.stack(data = list(y = as.vector(y)), A = A, effects = mesh.index)

```

We now create the model object. 

```{r}
rspde_model <- rspde.intrinsic.matern(mesh = mesh_2d, alpha = alpha)

```

Finally, we create the formula and fit the model to the data: 

```{r message=FALSE, warning=FALSE}
f <- y ~ -1 + f(field, model = rspde_model)
rspde_fit <- inla(f,
                  data = inla.stack.data(st.dat),
                  family = "gaussian",
                  control.predictor = list(A = inla.stack.A(st.dat)))

```

We can get a summary of the fit:
```{r}
summary(rspde_fit)
```

To get a summary of the fit of the random field only, we can do the following:
```{r}
result_fit <- rspde.result(rspde_fit, "field", rspde_model)
summary(result_fit)
tau <- op$tau
result_df <- data.frame(
  parameter = c("tau", "kappa"),
  true = c(tau, kappa), mean = c(result_fit$summary.tau$mean,
                                     result_fit$summary.kappa$mean),
  mode = c(result_fit$summary.tau$mode, result_fit$summary.kappa$mode)
)
print(result_df)
```

## Kriging with `R-INLA` implementation 

Let us now obtain predictions (i.e., do kriging) of the latent field on 
a dense grid in the region.

We begin by creating the grid of locations where we want to compute the predictions. To this end,
we can use the `rspde.mesh.projector()` function. This function has the same arguments
as the function `inla.mesh.projector()` the only difference being that the rSPDE
version also has an argument `nu` and an argument `rspde.order`. Thus, we
proceed in the same fashion as we would in `r inla_link()`'s standard SPDE implementation:


```{r create_proj_grid}
projgrid <- inla.mesh.projector(mesh_2d,
  xlim = c(0, 1),
  ylim = c(0, 1)
)
```

This lattice contains 100 × 100 locations (the default) which are shown in the following figure:


```{r plot_prd, fig.align='center'}
coord.prd <- projgrid$lattice$loc
plot(coord.prd, type = "p", cex = 0.1)
```

Let us now calculate the predictions jointly with the estimation. To this end, first, 
we begin by linking the prediction coordinates to the mesh nodes through an $A$
 matrix

```{r A_prd}
A.prd <- projgrid$proj$A
```

We now make a stack for the prediction locations. We have no data at the prediction locations, so we set `y=
NA`. We then join this stack with the estimation stack.

```{r stk.prd}
ef.prd <- list(c(mesh.index))
st.prd <- inla.stack(
  data = list(y = NA),
  A = list(A.prd), tag = "prd",
  effects = ef.prd
)
st.all <- inla.stack(st.dat, st.prd)
```

Doing the joint estimation takes a while, and we therefore turn off the computation of certain things that we are not interested in, such as the marginals for the random effect. 
We will also use a simplified integration strategy (actually only using the posterior mode of the hyper-parameters) through the command `control.inla = list(int.strategy = "eb")`, i.e. empirical Bayes:

```{r fit_prd, message=FALSE, warning=FALSE}
rspde_fitprd <- inla(f,
  family = "Gaussian",
  data = inla.stack.data(st.all),
  control.predictor = list(
    A = inla.stack.A(st.all),
    compute = TRUE, link = 1
  ),
  control.compute = list(
    return.marginals = FALSE,
    return.marginals.predictor = FALSE
  ),
  control.inla = list(int.strategy = "eb")
)
```

We then extract the indices to the prediction nodes and then extract the mean and the standard deviation of the response:

```{r stk.mean.sd}
id.prd <- inla.stack.index(st.all, "prd")$data
m.prd <- matrix(rspde_fitprd$summary.fitted.values$mean[id.prd], 100, 100)
sd.prd <- matrix(rspde_fitprd$summary.fitted.values$sd[id.prd], 100, 100)
```

Finally, we plot the results. First the mean:

```{r plot_pred, echo=TRUE, fig.align='center'}
field.pred.df <- data.frame(x1 = projgrid$lattice$loc[,1],
                        x2 = projgrid$lattice$loc[,2], 
                        y = as.vector(m.prd))
ggplot(field.pred.df, aes(x = x1, y = x2, fill = y)) +
  geom_raster()  + scale_fill_viridis()
```

Then, the marginal standard deviations:

```{r plot_pred_sd, fig.align='center', echo=TRUE}
field.pred.sd.df <- data.frame(x1 = proj$lattice$loc[,1],
                        x2 = proj$lattice$loc[,2], 
                        sd = as.vector(sd.prd))
ggplot(field.pred.sd.df, aes(x = x1, y = x2, fill = sd)) +
  geom_raster() + scale_fill_viridis()
```
