---
title: "inlabru implementation of the rational SPDE approach"
author: "David Bolin and Alexandre B. Simas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inlabru implementation of the rational SPDE approach}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: bolin19
  title: "The rational SPDE approach for Gaussian random fields with general smoothness"
  author:
  - family: Bolin
    given: David
  - family: Kirchner
    given: Kristin
  container-title: Journal of Computational and Graphical Statistics
  volume: 29
  issue: 2
  page: 274-285
  type: article-journal
  issued:
    year: 2020
- id: lindgren11
  title: "An explicit link between Gaussian fields and Gaussian Markov random fields: the stochastic partial differential equation approach"
  author:
  - family: Lindgren
    given: Finn
  - family: Rue
    given: Håvard
  - family: Lindström
    given: Johan
  container-title: Journal of the Royal Statistical Society. Series B. Statistical Methodology
  volume: 73
  issue: 4
  page: 423--498
  type: article-journal
  issued:
    year: 2011
- id: xiong22
  title: "Covariance-based rational approximations of fractional SPDEs for computationally efficient Bayesian inference"
  author: 
  - family: Xiong
    given: Zhen
  - family: Simas
    given: Alexandre B.
  - family: Bolin
    given: David
  container-title: arXiv:2209.04670
  type: preprint
  issued:
    year: 2022
- id: Hofreither21
  title: "An algorithm for best rational approximation based on barycentric rational interpolation"
  author:
  - family: Hofreither
    given: Clemens
  container-title: Numerical Algorithms
  volume: 88
  issue: 1
  page: 365--388
  type: article-journal
  issued:
    year: 2021 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1, kind = "Mersenne-Twister", normal.kind = "Inversion")
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction 

In this vignette we will present the [`inlabru`](http://inlabru.org/) implementation of the covariance-based
rational SPDE approach. For further technical details on the covariance-based approach, see 
the [Rational approximation with the `rSPDE` package](rspde_cov.html) vignette
and [@xiong22](https://arxiv.org/abs/2209.04670). 

We begin by providing a 
step-by-step illustration on how to use our implementation. To this end we 
will consider a real world data set that consists of precipitation measurements 
from the Paraná region in Brazil. 

After the initial model fitting, we will show how to change some parameters of the model. In the end, we will also provide an example in which we have replicates.

The examples in this vignette are the same as those in the [R-INLA implementation of the rational SPDE approach](rspde_inla.html) vignette. 
As in that case, it is important to mention that one can improve the performance
by using the PARDISO solver. Please, go to https://www.pardiso-project.org/r-inla/#license
to apply for a license. Also, use `inla.pardiso()` for instructions on 
how to enable the PARDISO sparse library.

## An example with real data

To illustrate our implementation of `rSPDE` in [`inlabru`](http://inlabru.org/) we will consider a dataset
available in `r inla_link()`. This data has also been used to illustrate the SPDE approach,
see for instance the book [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://www.routledge.com/Advanced-Spatial-Modeling-with-Stochastic-Partial-Differential-Equations/Krainski-Gomez-Rubio-Bakka-Lenzi-Castro-Camilo-Simpson-Lindgren-Rue/p/book/9780367570644) and also the vignette [Spatial Statistics using R-INLA and Gaussian Markov random fields](https://sites.stat.washington.edu/peter/591/INLA.html).
See also [@lindgren11](https://rss.onlinelibrary.wiley.com/doi/full/10.1111/j.1467-9868.2011.00777.x)
for theoretical details on the standard SPDE approach.

The data consist of precipitation measurements from the Paraná region in Brazil
and were provided by the Brazilian National Water Agency. The data were collected
at 616 gauge stations in Paraná state, south of Brazil, for each day in 2011.

### An rSPDE model for precipitation

We will follow the vignette [Spatial Statistics using R-INLA and Gaussian Markov random fields](https://sites.stat.washington.edu/peter/591/INLA.html). 
As precipitation data are always positive, we will assume it is Gamma distributed. 
`r inla_link()` uses the following parameterization of the Gamma distribution, 
$$\Gamma(\mu, \phi): \pi (y) = \frac{1}{\Gamma(\phi)} \left(\frac{\phi}{\mu}\right)^{\phi} y^{\phi - 1} \exp\left(-\frac{\phi y}{\mu}\right) .$$
In this parameterization, the distribution has expected value $E(x) = \mu$ and variance $V(x) = \mu^2/\phi$, where $1/\phi$ is a dispersion parameter.

In this example $\mu$ will be modelled using a stochastic model that includes both covariates 
and spatial structure, resulting in the latent Gaussian model for the precipitation measurements
$$\begin{align} y_i\mid \mu(s_i), \theta &\sim \Gamma(\mu(s_i),c\phi)\\ \log (\mu(s)) &= \eta(s) = \sum_k f_k(c_k(s))+u(s)\\ \theta &\sim \pi(\theta) \end{align},$$

where $y_i$ denotes the measurement taken at location $s_i$, $c_k(s)$ are covariates, $u(s)$ is a mean-zero Gaussian Matérn field, and $\theta$ is a vector containing all parameters of the model,
including smoothness of the field. That is, by using the `rSPDE` model we will 
also be able to estimate the smoothness of the latent field.

### Examining the data
We will be using [`inlabru`](http://inlabru.org/). The `inlabru` package is available on CRAN and 
also on [GitHub](https://github.com/inlabru-org/inlabru).

We begin by loading some libraries we need to get the data and build the plots.

```{r library_loading, message=FALSE}
library(gridExtra)
library(ggplot2)
library(lattice)
library(INLA)
library(inlabru)
library(splancs)
library(fields)
```

Let us load the data and the border of the region

```{r getting_data, message=FALSE}
data(PRprec)
data(PRborder)
```

The data frame contains daily measurements at 616 stations for the year 2011, as well as coordinates and altitude information for the measurement stations.
We will not analyze the full spatio-temporal data set, but instead look at the total precipitation in January, which we calculate as

```{r getting_january, message=FALSE}
Y <- rowMeans(PRprec[, 3 + 1:31])
```

In the next snippet of code, we extract the coordinates and altitudes 
and remove the locations with missing values.

```{r cleaning_data, message=FALSE}
ind <- !is.na(Y)
Y <- Y[ind]
coords <- as.matrix(PRprec[ind, 1:2])
alt <- PRprec$Altitude[ind]
```

Let us build a plot for the precipitations:

```{r plot_precipitations, message=FALSE, fig.align = "center", echo=TRUE}
p <- ggplot() +
  geom_point(aes(
    x = coords[, 1], y = coords[, 2],
    colour = Y
  ), size = 2, alpha = 1) +
  scale_colour_gradientn(colours = tim.colors(100)) +
  geom_path(aes(x = PRborder[, 1], y = PRborder[, 2])) +
  geom_path(aes(x = PRborder[1034:1078, 1], y = PRborder[
    1034:1078,
    2
  ]), colour = "red")
tryCatch(
  {
    print(p)
  },
  error = function(e) {
    print("Unable to build the plot")
  }
)
```

The red line in the figure shows the coast line, and we expect the distance to 
the coast to be a good covariate for precipitation. 

This covariate is not 
available, so let us calculate it for each observation location:

```{r getting_seaDist, message=FALSE}
seaDist <- apply(spDists(coords, PRborder[1034:1078, ],
  longlat = TRUE
), 1, min)
```

Now, let us plot the precipitation as a function of the possible covariates:

```{r plot_prec_as_func, message=FALSE, fig.width=7, fig.height=5, fig.align = "center"}
par(mfrow = c(2, 2))
plot(coords[, 1], Y, cex = 0.5, xlab = "Longitude")
plot(coords[, 2], Y, cex = 0.5, xlab = "Latitude")
plot(seaDist, Y, cex = 0.5, xlab = "Distance to sea")
plot(alt, Y, cex = 0.5, xlab = "Altitude")
par(mfrow = c(1, 1))
```


### Creating the rSPDE model

To use the [`inlabru`](http://inlabru.org/) implementation of the `rSPDE` model we need to load the functions:
```{r load_rspde_lib, message=FALSE}
library(rSPDE)
```

To create a `rSPDE` model, one would the `rspde.matern()` function in a similar fashion as one would
use the `inla.spde2.matern()` function. 

#### Mesh

We can use `r inla_link()` for creating the mesh. Let us create a mesh which is based 
on a non-convex hull to avoid adding many small triangles outside the domain 
of interest:

```{r mesh_creation, message=FALSE, fig.align='center'}
prdomain <- inla.nonconvex.hull(coords, -0.03, -0.05, resolution = c(100, 100))
prmesh <- inla.mesh.2d(boundary = prdomain, max.edge = c(0.45, 1), cutoff = 0.2)
plot(prmesh, asp = 1, main = "")
lines(PRborder, col = 3)
points(coords[, 1], coords[, 2], pch = 19, cex = 0.5, col = "red")
```


#### Setting up the data frame

In place of a `inla.stack`, we can set up a `data.frame()` to use [`inlabru`](http://inlabru.org/). We refer the
reader to vignettes in https://inlabru-org.github.io/inlabru/index.html for further details.

```{r}
prdata <- data.frame(long = coords[,1], lat = coords[,2], 
                        seaDist = inla.group(seaDist), y = Y)
coordinates(prdata) <- c("long","lat")
```

#### Setting up the rSPDE model

To set up an `rSPDE`model, all we need is the mesh. 
By default it will assume that we want to estimate the smoothness parameter $\nu$
and to do a covariance-based rational approximation of order 2.

Later in this vignette we will also see other options for setting up `rSPDE` models such as keeping
the smoothness parameter fixed and/or increasing the order of the covariance-based
rational approximation.

Therefore, to set up a model all we have to do is use the `rspde.matern()` function:

```{r create_model, message=FALSE}
rspde_model <- rspde.matern(mesh = prmesh)
```

Notice that this function is very reminiscent of `r inla_link()`'s `inla.spde2.matern()` function.

We will assume the following linkage between model components and observations 
$$\eta(s) \sim A x(s) + A \text{ Intercept} + \text{seaDist}.$$
$\eta(s)$ will then be used in the observation-likelihood, 
$$y_i\mid \eta(s_i),\theta \sim \Gamma(\exp(\eta (s_i)), c\phi).$$

### Model fitting

We will build a model using the distance to the sea $x_i$ 
as a covariate through an improper CAR(1) model with 
$\beta_{ij}=1(i\sim j)$, which `r inla_link()` calls a random walk of order 1.
We will fit it in `inlabru`'s  style:

```{r create_formula, message=FALSE}
cmp <- y ~ Intercept(1) + distSea(seaDist, model="rw1") +
field(coordinates, model = rspde_model)
```

To fit the model we simply use the `bru()` function:

```{r fit_model, message=FALSE, warning=FALSE}
rspde_fit <- bru(cmp, data = prdata,
  family = "Gamma",
  options = list(
    inla.mode = "experimental",
    control.inla = list(int.strategy = "eb"),
    verbose = FALSE)
)
```

### inlabru results

We can look at some summaries of the posterior distributions for the parameters, 
for example the fixed effects (i.e. the intercept) and the hyper-parameters 
(i.e. dispersion in the gamma likelihood, the precision of the RW1, 
and the parameters of the spatial field):

```{r get_summary}
summary(rspde_fit)
```

Let $\theta_1 = \textrm{Theta1}$, $\theta_2=\textrm{Theta2}$ and $\theta_3=\textrm{Theta3}$.
In terms of 
the SPDE
$$(\kappa^2 I - \Delta)^{\alpha/2}(\tau u) = \mathcal{W},$$
where $\alpha = \nu + d/2$, we have that
$$\tau = \exp(\theta_1),\quad \kappa = \exp(\theta_2), $$
and by default
$$\nu = 4\Big(\frac{\exp(\theta_3)}{1+\exp(\theta_3)}\Big).$$
The number 4 comes from the upper bound for $\nu$, which is discussed in [R-INLA implementation of the rational SPDE approach](rspde_inla.html) vignette. 

In general, we have
$$\nu = \nu_{UB}\Big(\frac{\exp(\theta_3)}{1+\exp(\theta_3)}\Big),$$
where $\nu_{UB}$ is the value of the upper bound for the smoothness parameter $\nu$.

Another choice for prior for $\nu$ is a truncated lognormal distribution and
is also discussed in [R-INLA implementation of the rational SPDE approach](rspde_inla.html) vignette.

### inlabru results in the original scale

We can obtain outputs with respect to parameters in the original scale by
using the function `rspde.result()`:

```{r get_result}
result_fit <- rspde.result(rspde_fit, "field", rspde_model)
summary(result_fit)
```

We can also plot the posterior densities. To this end we will use the `gg_df()` function, which creates `ggplot2` user-friendly data frames:

```{r plot_post, fig.align='center'}
posterior_df_fit <- gg_df(result_fit)

ggplot(posterior_df_fit) + geom_line(aes(x = x, y = y)) + 
facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```

### Predictions

Let us now obtain predictions (i.e. do kriging) of the expected precipitation on 
a dense grid in the region.

We begin by creating the grid in which we want to do the predictions. To this end,
we can use the `inla.mesh.projector()` function:


```{r create_proj_grid}
nxy <- c(150, 100)
projgrid <- inla.mesh.projector(prmesh,
  xlim = range(PRborder[, 1]),
  ylim = range(PRborder[, 2]), dims = nxy
)
```

This lattice contains 150 × 100 locations. 
One can easily change the resolution of the kriging prediction by changing `nxy`. 
Let us find the cells that are outside the region of interest so that we
do not plot the estimates there.

```{r get_inout}
xy.in <- inout(projgrid$lattice$loc, cbind(PRborder[, 1], PRborder[, 2]))
```

Let us plot the locations that we will do prediction:

```{r plot_prd, fig.align='center'}
coord.prd <- projgrid$lattice$loc[xy.in, ]
plot(coord.prd, type = "p", cex = 0.1)
lines(PRborder)
points(coords[, 1], coords[, 2], pch = 19, cex = 0.5, col = "red")
```

Let us now create a `data.frame()` of the coordinates:

```{r}
coord.prd.df <- data.frame(x1 = coord.prd[,1],
                            x2 = coord.prd[,2])
coordinates(coord.prd.df) <- c("x1", "x2")
```

Since we are using distance to the sea as a covariate, we also have to calculate this covariate for the prediction locations.
Finally, we add the prediction location to our prediction `data.frame()`, namely, `coord.prd.df`:

```{r pred_seaDist}
seaDist.prd <- apply(spDists(coord.prd,
  PRborder[1034:1078, ],
  longlat = TRUE
), 1, min)
coord.prd.df$seaDist <- seaDist.prd
```

```{r}
pred_obs <- predict(rspde_fit, coord.prd.df, 
        ~exp(Intercept + field + distSea))
```

Let us now build the data frame with the results:

```{r}
pred_df <- pred_obs@data
pred_df <- cbind(pred_df, pred_obs@coords)
```

Finally, we plot the results. First the predicted mean:

```{r, echo=TRUE, fig.align='center'}
p <- ggplot(pred_df, aes(x = x1, y = x2, fill = mean)) +
  geom_raster() +
  scale_fill_gradient(low = "yellow", high = "red")
p
```

Then, the std. deviations:

```{r plot_pred_sd_bru, fig.align='center', echo=TRUE}
p <- ggplot(pred_df, aes(x = x1, y = x2, fill = sd)) +
  geom_raster()
p
```

## An example with replicates

For this example we will simulate a data with replicates. We will
use the same example considered in the [Rational approximation with the `rSPDE` package](rspde_cov.html) vignette (the only difference is the way the data 
is organized).
We also refer the reader to this vignette for a description of the function
`matern.operators()`, along with its methods (for instance, the `simulate()` method).

### Simulating the data

Let us consider a simple Gaussian linear model with 30 independent replicates 
of a latent spatial field $x(\mathbf{s})$, observed at the same 
$m$ locations, $\{\mathbf{s}_1 , \ldots , \mathbf{s}_m \}$, for each replicate.
For each $i = 1,\ldots,m,$ we have

\begin{align} 
y_i &= x_1(\mathbf{s}_i)+\varepsilon_i,\\
\vdots &= \vdots\\

y_{i+29m} &= x_{30}(\mathbf{s}_i) + \varepsilon_{i+29m},
\end{align}

where $\varepsilon_1,\ldots,\varepsilon_{30m}$ are iid normally distributed
with mean 0 and standard deviation 0.1.

We use the basis function representation of $x(\cdot)$ to define the $A$ matrix
linking the point locations to the mesh. We also need to account for the fact
that we have 30 replicates at the same locations. To this end,
the $A$ matrix we need can be generated by `inla.spde.make.A()` function. The 
reason being that we are sampling $x(\cdot)$ directly and not the
latent vector described in the introduction of the 
[Rational approximation with the `rSPDE` package](rspde_cov.html) vignette.

We begin by creating the mesh:

```{r fig.align = "center"}
m <- 200
loc_2d_mesh <- matrix(runif(m * 2), m, 2)
mesh_2d <- inla.mesh.2d(
  loc = loc_2d_mesh,
  cutoff = 0.05,
  offset = c(0.1, 0.4),
  max.edge = c(0.05, 0.5)
)
plot(mesh_2d, main = "")
points(loc_2d_mesh[, 1], loc_2d_mesh[, 2])
```

We then compute the $A$ matrix, which is needed for simulation, and connects the observation
locations to the mesh:

```{r}
n.rep <- 30
A <- inla.spde.make.A(
  mesh = mesh_2d,
  loc = loc_2d_mesh,
  index = rep(1:m, times = n.rep),
  repl = rep(1:n.rep, each = m)
)
```

Notice that for the simulated data, we should use the $A$ matrix from `inla.spde.make.A()` function.

We will now simulate a latent process with standard deviation $\sigma=1$ and 
range $0.1$. We will use $\nu=0.5$ so that the model has an exponential covariance function.
To this end we create a model object with the `matern.operators()` function:

```{r}
nu <- 0.5
sigma <- 1
range <- 0.1
kappa <- sqrt(8 * nu) / range
tau <- sqrt(gamma(nu) / (sigma^2 * kappa^(2 * nu) * (4 * pi) * gamma(nu + 1)))
d <- 2
operator_information <- matern.operators(
  mesh = mesh_2d,
  nu = nu,
  kappa = kappa,
  sigma = sigma,
  m = 2
)
```
More details on this function can be found at the [Rational approximation with the rSPDE package](rspde_cov.html) vignette.

To simulate the latent process all we need to do is to use the `simulate()` method on
the `operator_information` object. We then obtain the simulated data $y$
by connecting with the $A$ matrix and adding the gaussian noise.

```{r}
set.seed(1)
u <- simulate(operator_information, nsim = n.rep)
y <- as.vector(A %*% as.vector(u)) +
  rnorm(m * n.rep) * 0.1
```

The first replicate of the simulated random field as well as the observation locations are shown in the following figure.

```{r, fig.show='hold', fig.align = "center",echo=TRUE}
opar <- par(mfrow = c(1, 2), mgp = c(1.2, 0.5, 0),
   mar = c(2, 2, 0.5, 0.5) + 0.1)
proj <- inla.mesh.projector(mesh_2d, dims = c(100, 100))
image(inla.mesh.project(proj, field = as.vector(u[, 1])),
  xlab = "", ylab = "",
  cex.main = 0.8, cex.axis = 0.8, cex.lab = 0.8
)
plot(loc_2d_mesh[, 1], loc_2d_mesh[, 2],
  cex = 0.2, pch = 16, xlab = "", ylab = "",
  cex.main = 0.8, cex.axis = 0.8, cex.lab = 0.8
)
par(opar)
```


### Fitting the inlabru rSPDE model

Let us then use the rational SPDE approach to fit the data.

We begin by creating the model object. 

```{r}
rspde_model.rep <- rspde.matern(mesh = mesh_2d,
          parameterization = "spde") 
```

Let us now create the `data.frame()` and the vector with the replicates indexes:

```{r}
rep.df <- data.frame(y = y, x1 = rep(loc_2d_mesh[,1], 30),
                      x2 = rep(loc_2d_mesh[,2], 30))
coordinates(rep.df) <- c("x1", "x2")
repl <- rep(1:30, each=200)
```

Let us create the component and fit. It is extremely important
not to forget the `replicate` when fitting model with 
the `bru()` function. It will not produce warning and might fit some
meaningless model.

```{r, message=FALSE, warning=FALSE}
cmp.rep <-
  y ~ -1 + field(coordinates,
    model = rspde_model.rep,
    replicate = repl
  )


rspde_fit.rep <-
  bru(cmp.rep,
    data = rep.df,
    family = "gaussian",
    options = list(
      list(inla.mode = "experimental")
    )
  )
```

We can get the summary:
```{r}
summary(rspde_fit.rep)
```

and the summary in the user's scale:
```{r}
result_fit_rep <- rspde.result(rspde_fit.rep, "field", rspde_model.rep)
summary(result_fit_rep)
result_df <- data.frame(
  parameter = c("tau", "kappa", "nu"),
  true = c(tau, kappa, nu),
  mean = c(
    result_fit_rep$summary.tau$mean,
    result_fit_rep$summary.kappa$mean,
    result_fit_rep$summary.nu$mean
  ),
  mode = c(
    result_fit_rep$summary.tau$mode,
    result_fit_rep$summary.kappa$mode,
    result_fit_rep$summary.nu$mode
  )
)
print(result_df)
```

## An example with a non-stationary model

Our goal now is to show how one can fit model with non-stationary $\sigma$ (std. deviation) and non-stationary $\rho$ (range parameter).
One can also use the parameterization in terms of non-stationary SPDE parameters $\kappa$ and $\tau$. 

We would like to mention that in the non-stationary case $\rho$ is an approximation of the range parameter.

For this example we will consider simulated data. 

### Simulating the data

Let us consider a simple Gaussian linear model with a latent spatial field $x(\mathbf{s})$,
defined on the rectangle $(0,10) \times (0,5)$, 
where the std. deviation and range parameter satisfy the following log-linear regressions:
\begin{align}
\log(\sigma(\mathbf{s})) &= \theta_1 + \theta_3 b(\mathbf{s}),\\
\log(\rho(\mathbf{s})) &= \theta_2 + \theta_3 b(\mathbf{s}),
\end{align}
where $b(\mathbf{s}) = (s_1-5)/10$. We assume the data is observed at 
$m$ locations, $\{\mathbf{s}_1 , \ldots , \mathbf{s}_m \}$.
For each $i = 1,\ldots,m,$ we have

$$y_i = x_1(\mathbf{s}_i)+\varepsilon_i,$$

where $\varepsilon_1,\ldots,\varepsilon_{m}$ are iid normally distributed
with mean 0 and standard deviation 0.1.

We begin by defining the domain and creating the mesh:

```{r fig.align = "center"}
rec_domain <- cbind(c(0, 1, 1, 0, 0) * 10, c(0, 0, 1, 1, 0) * 5)

mesh <- inla.mesh.2d(loc.domain = rec_domain, cutoff = 0.1, 
  max.edge = c(0.5, 1.5), offset = c(0.5, 1.5))
```

We follow the same structure as `INLA`. However, `INLA` only allows one to specify 
`B.tau` and `B.kappa` matrices, and, in `INLA`, if one wants to parameterize in terms of range and
standard deviation one needs to do it manually. Here we provide the option to directly provide
the matrices `B.sigma` and `B.range`. 

The usage of the matrices `B.tau` and `B.kappa` are identical to the corresponding ones in `inla.spde2.matern()` function.
The matrices `B.sigma` and `B.range` work in the same way, but they parameterize the stardard deviation
and range, respectively.

The columns of the `B` matrices correspond to the same parameter. The first column does not
have any parameter to be estimated, it is a constant column. 

So, for instance, if one wants to share a parameter with both `sigma` and `range` (or with both `tau` and `kappa`),
one simply let the corresponding column to be nonzero on both `B.sigma` and `B.range` (or on `B.tau` and `B.kappa`).

We will assume $\nu = 0.8$, $\theta_1 = 0, \theta_2 = 1$ and $\theta_3=1$. Let us now build the model with the `rspde.matern()` function:


```{r}
nu <- 0.8
true_theta <- c(0,1, 1)

# SPDE model
rspde_nonstat_sample <- rspde.matern(mesh, 
  B.sigma = cbind(0, 1, 0, (mesh$loc[,1] - 5) / 10), 
  B.range = cbind(0, 0, 1, (mesh$loc[,1] - 5) / 10),
  start.theta = true_theta,
  nu = nu)
```

In order to sample the data, we will obtain the precision matrix:

```{r}
prec_mat <- inla.cgeneric.q(rspde_nonstat_sample)$Q
```

Let us now sample the data with `inla.qsample()` function:

```{r, warning=FALSE}
u <- as.vector(inla.qsample(1, prec_mat, seed = 1))
```

Let us now obtain 400 random locations on the rectangle and 
compute the $A$ matrix:

```{r}
m <- 500
loc_mesh <- cbind(runif(m) * 10, runif(m) * 5)

Abar <- rspde.make.A(
  mesh = mesh,
  loc = loc_mesh
)
```

We can now generate the response vector `y`:
```{r}
y <- as.vector(Abar %*% as.vector(u)) + rnorm(m) * 0.1
```

### Fitting the inlabru rSPDE model

Let us then use the rational SPDE approach to fit the data.

We begin by creating the model object. We are creating a new one
so that we do not start the estimation at the true values.

```{r}
rspde_model_nonstat <- rspde.matern(mesh = mesh,
  B.sigma = cbind(0, 1, 0, (mesh$loc[,1] - 5) / 10), 
  B.range = cbind(0, 0, 1, (mesh$loc[,1] - 5) / 10)) 
```

Let us now create the `data.frame()` and the vector with the replicates indexes:

```{r}
nonstat_df <- data.frame(y = y, x1 = loc_mesh[,1],
                      x2 = loc_mesh[,2])
coordinates(nonstat_df) <- c("x1", "x2")
```

Let us create the component and fit. It is extremely important
not to forget the `replicate` when fitting model with 
the `bru()` function. It will not produce warning and might fit some
meaningless model.

```{r, message=FALSE, warning=FALSE}
cmp_nonstat <-
  y ~ -1 + field(coordinates,
    model = rspde_model_nonstat
  )


rspde_fit_nonstat <-
  bru(cmp_nonstat,
    data = nonstat_df,
    family = "gaussian",
    options = list(
      list(inla.mode = "experimental")
    )
  )
```

We can get the summary:
```{r}
summary(rspde_fit_nonstat)
```


We can obtain outputs with respect to parameters in the original scale by
using the function `rspde.result()`:

```{r get_result_nonstat}
result_fit_nonstat <- rspde.result(rspde_fit_nonstat, "field", rspde_model_nonstat)
summary(result_fit_nonstat)
```

Let us compare the mean to the true values of the parameters:

```{r}
summ_res_nonstat <- summary(result_fit_nonstat)
result_df <- data.frame(
  parameter = result_fit_nonstat$params,
  true = c(true_theta, nu),
  mean = summ_res_nonstat[,1],
  mode = summ_res_nonstat[,6]
)
print(result_df)
```

We can also plot the posterior densities. To this end we will use the `gg_df()` function, which creates `ggplot2` user-friendly data frames:

```{r plot_post_nonstat, fig.align='center'}
posterior_df_fit <- gg_df(result_fit_nonstat)

ggplot(posterior_df_fit) + geom_line(aes(x = x, y = y)) + 
facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```


## Further options of the `inlabru` implementation

There are several additional options that are available. For instance, it is possible to change the order of the rational approximation,
the upper bound for the smoothness parameter (which may speed up the fit), change the priors, change the type of the rational approximation,
among others. These options are described in the "Further options of the `rSPDE`-`INLA` implementation" 
section of the [R-INLA implementation of the rational SPDE approach](rspde_inla.html) vignette.
Observe that all these options are passed to the model through the `rspde.matern()` function, and therefore the resulting model
object can directly be used in the `bru()` function, in an identical manner to the examples above. 

## References 
