---
title: "An introduction to the rSPDE package"
author: "David Bolin and Alexandre B. Simas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to the rSPDE package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: bolin19
  title: "The rational SPDE approach for Gaussian random fields with general smoothness"
  author:
  - family: Bolin
    given: David
  - family: Kirchner
    given: Kristin
  container-title: Journal of Computational and Graphical Statistics
  volume: 29
  issue: 2
  page: 274-285
  type: article-journal
  issued:
    year: 2020
- id: lindgren11
  title: "An explicit link between Gaussian fields and Gaussian Markov random fields: the stochastic partial differential equation approach"
  author:
  - family: Lindgren
    given: Finn
  - family: Rue
    given: Håvard
  - family: Lindström
    given: Johan
  container-title: Journal of the Royal Statistical Society. Series B. Statistical Methodology
  volume: 73
  issue: 4
  page: 423--498
  type: article-journal
  issued:
    year: 2011
- id: bolin13
  title: "Spatial Matérn Fields Driven by Non-Gaussian Noise"
  author:
    - family: Bolin
      given: David
  container-title: Scandinavian Journal of Statistics
  volume: 41
  issue: 3
  page: 557-579
  type: article-journal
  issued:
    year: 2013
- id: wallin15
  title: "Geostatistical Modelling Using Non-Gaussian Matérn Fields"
  author:
    - family: Wallin
      given: Jonas
    - family: Bolin
      given: David
  container-title: Scandinavian Journal of Statistics
  volume: 42
  issue: 3
  page: 872-890
  type: article-journal
  issued:
    year: 2015
- id: asar20
  title: "Linear mixed effects models for non‐Gaussian repeated measurement data"
  author:
    - family: Asar
      given: Özgür
    - family: Bolin
      given: David
    - family: Diggle
      given: Peter
    - family: Wallin
      given: Jonas
  container-title: Journal of the Royal Statistical Society. Series C. Applied Statistics
  volume: 69
  issue: 5
  page: 1015-1065
  type: article-journal
  issued:
    year: 2020 
- id: xiong22
  title: "Covariance-based rational approximations of fractional SPDEs for computationally efficient Bayesian inference"
  author: 
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Xiong
    given: Zhen    
  container-title: Journal of Computational and Graphical Statistics
  type: article-journal
  issued:
    year: 2023   
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

set.seed(1)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction

In this vignette we provide a brief introduction to the `rSPDE`
package. The main approach for constructing the rational approximations is the covariance-based rational SPDE approach of [@xiong22](https://arxiv.org/abs/2209.04670). The package contains three main "families" of functions that implement the approach:

* an interface to `r inla_link()`;

* an interface to [`inlabru`](http://inlabru.org/);

* a stand-alone implementation of the approach.

To illustrate these different functions, we begin by using the package to generate a simple data set, which then will be analyzed using 
the different approaches. Further details on each family of functions is given in the following additional vignettes:

* [R-INLA implementation of the rational SPDE approach](rspde_inla.html)

* [inlabru implementation of the rational SPDE approach](rspde_inlabru.html)

* [Rational approximation with the rSPDE package](rspde_cov.html)

The `rSPDE` package also has a separate group of functions for performing the operator-based
rational approximations introduced in [@bolin19](https://www.tandfonline.com/doi/full/10.1080/10618600.2019.1665537). 
These are especially useful when performing rational 
approximations for fractional SPDE models with non-Gaussian noise. An
example in which such approximation is suitable is when one has the
so-called type-G Lévy noises. 

We refer the reader to [@wallin15](https://onlinelibrary.wiley.com/doi/full/10.1111/sjos.12141), 
[@bolin13](https://onlinelibrary.wiley.com/doi/abs/10.1111/sjos.12046) and
[@asar20](https://rss.onlinelibrary.wiley.com/doi/pdf/10.1111/rssc.12405)
for examples of models driven by type-G Lévy noises. We also refer the reader
to the [`ngme` package](https://github.com/davidbolin/ngme) where one can fit such models.

We explore the functions for performing the operator-based rational approximation
on the vignette:

* [Operator-based rational approximation with the rSPDE package](rspde_base.html)

## A toy data set

We begin by generating a toy data set.

For this illustration, we will simulate a data set on a two-dimensional spatial domain. 
To this end, we need to construct a mesh over the domain of interest and then compute 
the matrices needed to define the operator. We will use the `r inla_link()` package
to create the mesh and obtain the matrices of interest.

We will begin by defining a mesh over $[0,1]\times [0, 1]$:

```{r, message=FALSE, warning=FALSE, fig.align='center'}
library(INLA)
n_loc <- 2000
loc_2d_mesh <- matrix(runif(n_loc * 2), n_loc, 2)
mesh_2d <- inla.mesh.2d(
  loc = loc_2d_mesh,
  cutoff = 0.05,
  max.edge = c(0.1, 0.5)
)
plot(mesh_2d, main = "")
```

We now use the `matern.operators()` function to construct a rational SPDE approximation of 
order $m=2$ for a Gaussian random field with a Matérn covariance function on
$[0,1]\times [0, 1]$. We choose $\nu=0.5$ which corresponds to exponential
covariance. We also set $\sigma=1$ and the range as $0.2$.

```{r, message=FALSE}
library(rSPDE)
sigma <- 2
range <- 0.25
nu <- 1.3
kappa <- sqrt(8 * nu) / range
op <- matern.operators(
  mesh = mesh_2d, nu = nu,
  range = range, sigma = sigma, m = 2,
  parameterization = "matern"
)
tau <- op$tau
```

We can now use the `simulate` function to simulate a realization of the field $u$:

```{r}
u <- simulate(op)
```

Let us now consider a simple Gaussian linear model where 
the spatial field $u(\mathbf{s})$ is observed at
$m$ locations, $\{\mathbf{s}_1 , \ldots , \mathbf{s}_m \}$ under Gaussian measurement noise.
For each $i = 1,\ldots,m,$ we have
$$
\begin{align} 
y_i &= u(\mathbf{s}_i)+\varepsilon_i\\
\end{align},
$$
where $\varepsilon_1,\ldots,\varepsilon_{m}$ are iid normally distributed
with mean 0 and standard deviation 0.1.

To generate a data set `y` from this model, we first draw some observation locations at random in the domain and then
use the `r inla_link()` function `inla.spde.make.A()` to construct the observation matrix which can be used to evaluate the
simulated field $u$ at the observation locations. After this we simply add the measurment noise. 

```{r}
A <- inla.spde.make.A(
  mesh = mesh_2d,
  loc = loc_2d_mesh
)
sigma.e <- 0.1
y <- A %*% u + rnorm(n_loc) * sigma.e
```

The generated data can be seen in the following image.

```{r,fig.align = "center", echo=TRUE}
library(ggplot2)
library(viridis)
df <- data.frame(x1 = as.double(loc_2d_mesh[, 1]),
  x2 = as.double(loc_2d_mesh[, 2]), y = as.double(y))
ggplot(df, aes(x = x1, y = x2, col = y)) +
  geom_point() +
  scale_color_viridis()
```

The simulated random field is shown in the following figure. 

```{r, fig.show='hold', fig.align = "center",echo=TRUE, message=FALSE,warning=FALSE}
proj <- inla.mesh.projector(mesh_2d, dims = c(100, 100))
field <- inla.mesh.project(proj, field = as.vector(u))
field.df <- data.frame(x1 = proj$lattice$loc[,1],
                        x2 = proj$lattice$loc[,2], 
                        y = as.vector(field))
ggplot(field.df, aes(x = x1, y = x2, fill = y)) +
  geom_raster() + xlim(0,1) + ylim(0,1) + 
  scale_fill_viridis()
```

## Fitting the model with `R-INLA` implementation of the rational SPDE approach

We will now fit the model of the toy data set using our `r inla_link()` implementation of the
rational SPDE approach. Further details on this implementation can be found in
 [R-INLA implementation of the rational SPDE approach](rspde_inla.html).
 
We begin by creating the $A$ matrix, the index, and the
`inla.stack` object.

```{r}
Abar <- rspde.make.A(mesh = mesh_2d, loc = loc_2d_mesh)
mesh.index <- rspde.make.index(name = "field", mesh = mesh_2d)

st.dat <- inla.stack(
  data = list(y = as.vector(y)),
  A = Abar,
  effects = mesh.index
)
```

We now create the model object. We need to set
an upper bound for the smoothness parameter $\nu$. The default value for this is $4$. If 
we increase the upper bound for $\nu$
we also increase the computational cost, and if we decrease the upper bound we also decrease the computatoinal cost. 
For this example we set `nu.upper.bound=2`.
See the  [R-INLA implementation of the rational SPDE approach](rspde_inla.html)
for further details.

```{r}
rspde_model <- rspde.matern(
  mesh = mesh_2d,
  nu.upper.bound = 2,
  parameterization = "spde"
)
```

Finally, we create the formula and fit the model to the data: 

```{r message=FALSE, warning=FALSE}
f <-
  y ~ -1 + f(field, model = rspde_model)
rspde_fit <-
  inla(f,
    data = inla.stack.data(st.dat),
    family = "gaussian",
    control.predictor =
      list(A = inla.stack.A(st.dat))
  )
```

We can get a summary of the fit:
```{r}
summary(rspde_fit)
```

To get a summary of the fit of the random field only, we can do the following:
```{r}
result_fit <- rspde.result(rspde_fit, "field", rspde_model)
summary(result_fit)
tau <- op$tau
result_df <- data.frame(
  parameter = c("tau", "kappa", "nu"),
  true = c(tau, kappa, nu), mean = c(
    result_fit$summary.tau$mean,
    result_fit$summary.kappa$mean,
    result_fit$summary.nu$mean
  ),
  mode = c(
    result_fit$summary.tau$mode,
    result_fit$summary.kappa$mode,
    result_fit$summary.nu$mode
  )
)
print(result_df)
```

We can also obtain the summary in the `matern` parameterization by setting the `parameterization` argument to `matern`:

```{r}
result_fit_matern <- rspde.result(rspde_fit, "field", rspde_model,
                                  parameterization = "matern")
summary(result_fit_matern)
result_df_matern <- data.frame(
  parameter = c("sigma", "range", "nu"),
  true = c(sigma, range, nu), mean = c(
    result_fit_matern$summary.std.dev$mean,
    result_fit_matern$summary.range$mean,
    result_fit_matern$summary.nu$mean
  ),
  mode = c(
    result_fit_matern$summary.std.dev$mode,
    result_fit_matern$summary.range$mode,
    result_fit_matern$summary.nu$mode
  )
)
print(result_df_matern)
```

## Kriging with `R-INLA` implementation of the rational SPDE approach

Let us now obtain predictions (i.e., do kriging) of the latent field on 
a dense grid in the region.

We begin by creating the grid of locations where we want to compute the predictions. To this end,
we can use the `rspde.mesh.projector()` function. This function has the same arguments
as the function `inla.mesh.projector()` the only difference being that the rSPDE
version also has an argument `nu` and an argument `rspde.order`. Thus, we
proceed in the same fashion as we would in `r inla_link()`'s standard SPDE implementation:


```{r create_proj_grid}
projgrid <- rspde.mesh.projector(mesh_2d,
  xlim = c(0, 1),
  ylim = c(0, 1)
)
```

This lattice contains 100 × 100 locations (the default) which are shown in the following figure:


```{r plot_prd, fig.align='center'}
coord.prd <- projgrid$lattice$loc
plot(coord.prd, type = "p", cex = 0.1)
```

Let us now calculate the predictions jointly with the estimation. To this end, first, 
we begin by linking the prediction coordinates to the mesh nodes through an $A$
 matrix

```{r A_prd}
A.prd <- projgrid$proj$A
```

We now make a stack for the prediction locations. We have no data at the prediction locations, so we set `y=
NA`. We then join this stack with the estimation stack.

```{r stk.prd}
ef.prd <- list(c(mesh.index))
st.prd <- inla.stack(
  data = list(y = NA),
  A = list(A.prd), tag = "prd",
  effects = ef.prd
)
st.all <- inla.stack(st.dat, st.prd)
```

Doing the joint estimation takes a while, and we therefore turn off the computation of certain things that we are not interested in, such as the marginals for the random effect. 
We will also use a simplified integration strategy (actually only using the posterior mode of the hyper-parameters) through the command `control.inla = list(int.strategy = "eb")`, i.e. empirical Bayes:

```{r fit_prd, message=FALSE, warning=FALSE}
rspde_fitprd <- inla(f,
  family = "Gaussian",
  data = inla.stack.data(st.all),
  control.predictor = list(
    A = inla.stack.A(st.all),
    compute = TRUE, link = 1
  ),
  control.compute = list(
    return.marginals = FALSE,
    return.marginals.predictor = FALSE
  ),
  control.inla = list(int.strategy = "eb")
)
```

We then extract the indices to the prediction nodes and then extract the mean and the standard deviation of the response:

```{r stk.mean.sd}
id.prd <- inla.stack.index(st.all, "prd")$data
m.prd <- matrix(rspde_fitprd$summary.fitted.values$mean[id.prd], 100, 100)
sd.prd <- matrix(rspde_fitprd$summary.fitted.values$sd[id.prd], 100, 100)
```

Finally, we plot the results. First the mean:

```{r plot_pred, echo=TRUE, fig.align='center'}
field.pred.df <- data.frame(x1 = projgrid$lattice$loc[,1],
                        x2 = projgrid$lattice$loc[,2], 
                        y = as.vector(m.prd))
ggplot(field.pred.df, aes(x = x1, y = x2, fill = y)) +
  geom_raster() + xlim(0,1) + ylim(0,1) + 
  scale_fill_viridis()
```

Then, the marginal standard deviations:

```{r plot_pred_sd, fig.align='center', echo=TRUE}
field.pred.sd.df <- data.frame(x1 = proj$lattice$loc[,1],
                        x2 = proj$lattice$loc[,2], 
                        sd = as.vector(sd.prd))
ggplot(field.pred.sd.df, aes(x = x1, y = x2, fill = sd)) +
  geom_raster() + xlim(0,1) + ylim(0,1) + 
  geom_raster() +
  scale_fill_viridis()
```



## Fitting the model with `inlabru` implementation of the rational SPDE approach

We will now fit the same model of the toy data set using our [`inlabru`](http://inlabru.org/) implementation of the
rational SPDE approach. Further details on this implementation can be found in
 [`inlabru` implementation of the rational SPDE approach](rspde_inlabru.html).

 We begin by loading the `inlabru` package:

 ```{r message=FALSE}
 library(inlabru)
 ```
 
The creation of the model object is the same as in `R-INLA`'s case:

```{r}
rspde_model <- rspde.matern(
  mesh = mesh_2d,
  nu.upper.bound = 2,
  parameterization = "spde"
)
```

The advantage with `inlabru` is that we do not need to form the stack manually, but can simply collect the required data in 
a `data.frame()`:

```{r}
toy_df <- data.frame(coord1 = loc_2d_mesh[,1],
                     coord2 = loc_2d_mesh[,2],
                     y = as.vector(y))
coordinates(toy_df) <- c("coord1", "coord2")
```

Finally, we create the component and fit:

```{r message=FALSE, warning=FALSE}
cmp <-
  y ~ -1 + field(coordinates, model = rspde_model)

rspde_bru_fit <-
  bru(cmp,
      data=toy_df,
    options=list(
    family = "gaussian")
  )
```

At this stage, we can get a summary of the fit just as in the `R-INLA` case:
```{r}
summary(rspde_bru_fit)
```

and also obtain a summary of the field only:
```{r}
result_fit <- rspde.result(rspde_bru_fit, "field", rspde_model)
summary(result_fit)
tau <- op$tau
result_df <- data.frame(
  parameter = c("tau", "kappa", "nu"),
  true = c(tau, kappa, nu), mean = c(
    result_fit$summary.tau$mean,
    result_fit$summary.kappa$mean,
    result_fit$summary.nu$mean
  ),
  mode = c(
    result_fit$summary.tau$mode,
    result_fit$summary.kappa$mode,
    result_fit$summary.nu$mode
  )
)
print(result_df)
```


Let us obtain a summary in the `matern` parameterization by setting the `parameterization` argument to `matern`:

```{r}
result_fit_matern <- rspde.result(rspde_bru_fit, "field", rspde_model,
                                  parameterization = "matern")
summary(result_fit_matern)
result_df_matern <- data.frame(
  parameter = c("sigma", "range", "nu"),
  true = c(sigma, range, nu), mean = c(
    result_fit_matern$summary.std.dev$mean,
    result_fit_matern$summary.range$mean,
    result_fit_matern$summary.nu$mean
  ),
  mode = c(
    result_fit_matern$summary.std.dev$mode,
    result_fit_matern$summary.range$mode,
    result_fit_matern$summary.nu$mode
  )
)
print(result_df_matern)
```

## Kriging with `inlabru` implementation of the rational SPDE approach

Let us now obtain predictions (i.e., do kriging) of the latent field on 
a dense grid in the region.

We begin by creating the grid of the locations where we want to evaluate the predictions. We begin
by creating a regular grid in and then extract the coorinates:
```{r}
pred_coords <- data.frame(x1 = projgrid$lattice$loc[,1],
                          x2 = projgrid$lattice$loc[,2])
coordinates(pred_coords) <- c("x1", "x2")
```

Let us now compute the predictions. An advantage with `inlabru` is that we can do this after fitting the model to the data:

```{r}
field_pred <- predict(rspde_bru_fit, pred_coords, ~field)
```

The following figure shows the mean of these predictions:

```{r, echo=TRUE, fig.align='center', message=FALSE, warning=FALSE}
field_pred_df <- field_pred@data
field_pred_df <- cbind(field_pred_df, field_pred@coords)
ggplot(field_pred_df, aes(x = x1, y = x2, fill = mean)) +
  geom_raster() + xlim(0,1) + ylim(0,1) + 
  scale_fill_viridis()
```

The following figure shows the marginal standard deviations of the predictions:

```{r plot_pred_sd_bru, fig.align='center', echo=TRUE, message=FALSE, warning=FALSE}
ggplot(field_pred_df, aes(x = x1, y = x2, fill = sd)) +
  geom_raster() + xlim(0,1) + ylim(0,1) + 
  scale_fill_viridis()
```

An alternative and very simple approach is to use the `pixels()` function:

```{r plot_pred_pxl, fig.align='center', warning=FALSE}
pxl <- pixels(mesh_2d)

field_pred <- predict(rspde_bru_fit, pxl, ~field)

ggplot() + gg(field_pred) +
  scale_fill_viridis() + xlim(0,1) + ylim(0,1)
```

## Fitting the model with `rSPDE`

We will now fit the model of the toy data set without using `r inla_link()` or `inlabru`.
To this end we will use the rational approximation functions from 
`rSPDE` package. Further details can be found in the vignette
[Rational approximation with the rSPDE package](rspde_cov.html).

We use the 
function `rSPDE.construct.matern.loglike()` to define the likelihood. This function
is object-based, in the sense that it obtains several of the quantities
it needs from the `rSPDE` model object. 

Notice that we already created a `rSPDE` model object to simulate the data.
We will, then, use the same model object. Recall that the `rSPDE` model
object we created is `op`. 

Let us create an object for estimation, a `data.frame` with the data and then fit the model using the `rspde_lme()` function.

```{r}
op_est <- matern.operators(
  mesh = mesh_2d, m = 2
)

toy_df_rspde <- data.frame(coord1 = loc_2d_mesh[,1],
                     coord2 = loc_2d_mesh[,2],
                     y = as.vector(y))
```

```{r}
fit_rspde <- rspde_lme(y ~ -1, data = toy_df_rspde, loc = c("coord1", "coord2"),
                      model = op_est, parallel = TRUE)
```

We can obtain the summary:

```{r}
summary(fit_rspde)
```

Let us compare with the true values:
```{r}
print(data.frame(
  sigma = c(sigma, fit_rspde$matern_coeff$random_effects[2]), 
  range = c(range, fit_rspde$matern_coeff$random_effects[3]),
  nu = c(nu, fit_rspde$matern_coeff$random_effects[1]),
  row.names = c("Truth", "Estimates")
))

# Time to fit
print(fit_rspde$fitting_tim)
```


## Kriging with `rSPDE`

We will now do kriging on the same dense grid
we did for the `r inla_link()`-based rational SPDE approach,
but now using the `rSPDE` functions. To this end we
will use the `predict` method on the `rSPDE` model object.

Observe that we need an $A$ matrix connecting the mesh to the 
prediction locations. 

Let us now create the `data.frame` with the prediction locations:

```{r create_proj_grid_prd2}
predgrid <- inla.mesh.projector(mesh_2d,
  xlim = c(0, 1),
  ylim = c(0, 1)
)
pred_coords <- data.frame(coord1 = predgrid$lattice$loc[,1],
                          coord2 = predgrid$lattice$loc[,2])
```

We will now use the `predict()` method on the `rSPDE` model object
with the argument `compute.variances` set to `TRUE` so that
we can plot the standard deviations. Let us also update the values of 
the `rSPDE` model object to the fitted ones, and also save the estimated
value of `sigma.e`.

```{r}
pred.rspde <- predict(fit_rspde,
  data = pred_coords, loc = c("coord1", "coord2"),
  compute_variances = TRUE
)
```

Finally, we plot the results. First the mean:

```{r plot_pred2, fig.align='center', echo=TRUE}
field.pred2.df <- data.frame(x1 = predgrid$lattice$loc[,1],
                             x2 = predgrid$lattice$loc[,2],
                             y = as.vector(pred.rspde$mean))
ggplot(field.pred2.df, aes(x = x1, y = x2, fill = y)) +
  geom_raster() + xlim(0,1) + ylim(0,1) + 
  scale_fill_viridis()
```

Then, the standard deviations:

```{r plot_pred_sd2, fig.align='center', echo=TRUE}
field.pred2.sd.df <-field.pred2.df <- data.frame(x1 = predgrid$lattice$loc[,1],
                             x2 = predgrid$lattice$loc[,2],
                             sd = as.vector(sqrt(pred.rspde$variance)))
ggplot(field.pred2.sd.df, aes(x = x1, y = x2, fill = sd)) +
  geom_raster() +
  scale_fill_viridis()
```


# References

