---
title: "Spatio-temporal models"
author: "David Bolin and Alexandre B. Simas"
date: "Created: 2024-10-23. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: lindgren24
  title: "A diffusion-based spatio-temporal extension of Gaussian Matérn fields"
  author:
  - family: Lindgren
    given: Finn
  - family: Bakka
    given: Haakon  
  - family: Bolin
    given: David
  - family: Krainski
    given: Elias
  - family: Rue
    given: Håvard
  container-title: SORT statistics and operations research transactions
  volume: 48
  issue: 1
  page: 3-66
  type: article-journal
  issued:
    year: 2024
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rSPDE)
set.seed(12345)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction

The `rSPDE` package implements the following spatio-temporal model 
$$
d u + \gamma(\kappa^2 + \kappa^{d/2}\rho\cdot \nabla - \Delta)^{\alpha} u = dW_Q, \quad \text{on } T\times D
$$
where $T$ is a temporal interval and $D$ is a spatial domain which can be an interval, a bounded subset of $\mathbb{R}^2$ or a metric graph. Here $\kappa>0$ is a spatial range parameter, $\rho$ is a drift parameter which is in $\mathbb{R}$ for spatial domains that are intervals or metric graphs, and in $\mathbb{R}^2$ for spatial domains which are bounded subsets of $\mathbb{R}^2$. Further, $W_Q$ is a $Q$-Wiener process with spatial covariance operator $\sigma^2(\kappa^2 - \Delta)^{-\beta}$, where $\sigma^2$ is a variance parameter. Thus, the model has two smoothness parameters $\alpha$ and $\beta$ which are assumed to be integers. The model is therefore a generalization of the spatio-temporal models introduced in 
[@lindgren24](https://www.raco.cat/index.php/SORT/article/view/428665), where the generalization is to allow for drift and to allow for metric graphs as spatial domains. The model is implemented using a finite element discretization of the corresponding precision operator 
$$
\sigma^{-2}(d + \gamma(\kappa^2 + \kappa^{d/2}\rho\cdot \nabla - \Delta)^{\alpha})((d + \gamma(\kappa^2 - \kappa^{d/2}\rho\cdot \nabla - \Delta)^{\alpha}))(\kappa^2 - \rho\cdot \nabla - \Delta)^{\beta}
$$
in both space and time, similarly to the discretization introduced in [@lindgren24](https://www.raco.cat/index.php/SORT/article/view/428665). 


## Implementation details 
Let us begin by loading some packages needed for making the plots
```{r library_loading, message=FALSE}
library(ggplot2)
library(gridExtra)
library(viridis)
```
The function `spacetime.operators()` can be used to define the model. The function requires specifying the two smoothness parameters, and the discretization points for the spatial and temporal discretizations. The spatial discretization can be specified through a mesh object from the `fmesher` package, as a graph from the `MetricGraph` package, or as the mesh nodes for models on intervals. The temporal discretization can be specified either by specifying the mesh nodes or by providing a mesh object. 

Assume that we want to define a model on the spatial interval $[0,20]$ and the temporal domain $[0,10]$. We can then simply specify the mesh nodes as  
```{r}
s <- seq(from = 0, to = 20, length.out = 101)
t <- seq(from = 0, to = 10, length.out = 21)
```
We can now use `spacetime.operators()` to construct the model
```{r}
kappa <- 5
sigma <- 10
gamma <- 1/20
rho_1d <- 1
alpha <- 1
beta <- 1

op <- spacetime.operators(space_loc = s, time_loc = t,
                          kappa = kappa, sigma = sigma, alpha = alpha,
                          beta = beta, rho = rho_1d, gamma = gamma)
```

The `spacetime.operators` object has a `plot_covariances` method which for univariate spatial domains simply plots the covariance 
$C(u(s,t), u(s_0, t_0))$ for a fixed spatio-temporal location $(s_0, t_0)$ specified by the indices in the spatial and temporal discretizations. For example:
```{r}
op$plot_covariances(t.ind = 15, s.ind = 50)
```
The object `op` contains the matrices needed for evaluating the model, and we have here initialized it by providing values for all parameters. 

We can simulate from the model using `simulate()`: 

```{r}
u <- simulate(op)
```

There is also built-in support for kriging prediction. To illustrate this, we use the simulation to create some noisy observations of the process. For this, we first construct the observation matrix linking the FEM basis functions to the locations where we want to simulate. We first randomly generate some observation locations and then construct the matrix.
```{r}
n.obs <- 500

obs.loc <- data.frame(x = max(s)*runif(n.obs), t = max(t)*runif(n.obs))
A <- op$make_A(obs.loc$x, obs.loc$t)
```

We now generate the observations as $Y_i = u(s_i) + \varepsilon_i$, where $\varepsilon_i \sim N(0,\sigma_e^2)$ is Gaussian measurement noise. 
```{r}
x <- simulate(op, nsim = 1) 
sigma.e <- 0.01
Y <- as.vector(A%*%x + sigma.e*rnorm(n.obs))
```

Finally, we compute the kriging prediction of the process $u$ at the locations in `s` based on these observations. To specify which locations that should be predicted, the argument `Aprd` is used. This argument should be an observation matrix that links the mesh locations to the prediction locations.  
```{r, fig.show='hold',fig.align = "center"}
Aprd <- op$make_A(rep(s, length(t)), rep(t, each = length(s)))
u.krig <- predict(op, A = A, Aprd = Aprd, Y = Y, sigma.e = sigma.e)
```
The process simulation,  and the kriging prediction are shown in the following figure.

```{r, fig.show='hold', fig.align = "center",echo=TRUE}
data.df <- data.frame(space = obs.loc$x, time = obs.loc$t, field = Y, type = "Data")
krig.df <- data.frame(space = rep(s, length(t)), time = rep(t, each = length(s)),
                      field = as.vector(u.krig$mean), type = "Prediction")
df_plot <- rbind(data.df, krig.df)
ggplot(df_plot) + aes(x = space, y = time, fill = field) +
        facet_wrap(~type) + geom_raster(data = krig.df) +
        geom_point(data = data.df, aes(colour = field),
        show.legend = FALSE) + 
        scale_fill_viridis() + scale_colour_viridis()

```

## Parameter estimation

To estimate the model parameters based on this data, we can use the our `rspde_lme` function or our `inlabru` implementation. For this, we collect the data in a data frame, that also contanis the spatial locations, and

```{r}
df_1d <- data.frame(y = as.matrix(Y), space = obs.loc$x, time = obs.loc$t)
```

### `rspde_lme` implementation
We now fit the model:

```{r}
res <- rspde_lme(y ~ 1, loc = "space", loc_time = "time", data = df_1d, model = op, parallel = TRUE)

```

In the call, `y~1` indicates that we also want to estimate a mean value of the model, and the arguments `loc` and `loc_time` provides the names of the spatial and temporal coordinates in the data frame. 
Let us see a summary of the fitted model:

```{r}
summary(res)
```

Let us compare the estimated results with the true values:
```{r}
results <- data.frame(kappa = c(kappa, res$coeff$random_effects[1]), 
                      sigma = c(sigma, res$coeff$random_effects[2]),
                      gamma = c(gamma, res$coeff$random_effects[3]), 
                      rho = c(rho_1d, res$coeff$random_effects[4]),
                      sigma.e = c(sigma.e, res$coeff$measurement_error),
                      intercept = c(0, res$coeff$fixed_effects),
                      row.names = c("True", "Estimate"))
                      
print(results)
```
Finally, we can also do prediction based on the fitted model as 
```{r}
pred.data <- data.frame(x = rep(s,length(t)), t = rep(t,each=length(s)))
pred <- predict(res, newdata = pred.data, loc = "x", time = "t")
data.df <- data.frame(space = obs.loc$x, time = obs.loc$t, field = Y, type = "Data")
krig.df <- data.frame(space = rep(s, length(t)), time = rep(t, each = length(s)),
                      field = as.vector(pred$mean), type = "Prediction")
df_plot <- rbind(data.df, krig.df)
ggplot(df_plot) + aes(x = space, y = time, fill = field) +
        facet_wrap(~type) + geom_raster(data = krig.df) +
        geom_point(data = data.df, aes(colour = field),
        show.legend = FALSE) + 
        scale_fill_viridis() + scale_colour_viridis()

```

### `inlabru` Implementation

Let us now fit the model using our `inlabru` implementation. We start by creating the model object with the `rspde.spacetime()` function:
```{r}
library(inlabru)
st_bru <- rspde.spacetime(space_loc = s, time_loc = t, alpha = 1, beta=1)
```

We now create the model component, which requires the user to pass the index as a list containing the elements `space` with the spatial indices and `time` with the temporal indices:

```{r}
cmp <- y ~ -1 + Intercept(1) + field(list(space=space, time = time), model = st_bru)
```

We are now in a position to fit the model:

```{r}
bru_fit <- bru(cmp, data = df_1d)
```

Let us now compare the estimated results (the means of the parameters) with the true values:

```{r}
param_st_1d <- transform_parameters_spacetime(
                        bru_fit$summary.hyperpar$mean[2:5], 
                                            st_bru)

results <- data.frame(kappa = c(kappa,  param_st_1d[[1]]), 
                      sigma = c(sigma,  param_st_1d[[2]]),
                      gamma = c(gamma,  param_st_1d[[3]]), 
                      rho = c(rho_1d, param_st_1d[[4]]),
                      sigma.e = c(sigma.e, sqrt(1/bru_fit$summary.hyperpar$mean[1])),
                      intercept = c(0, bru_fit$summary.fixed$mean),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

## A spatial example
Let us now illustrate how to implement a spatial version. We start by creating a 
region of interest and a spatial mesh using the `fmesher` package:
```{r}
library(fmesher)
n_loc <- 1000
loc_2d_mesh <- matrix(runif(n_loc * 2), n_loc, 2)
mesh_2d <- fm_mesh_2d(
  loc = loc_2d_mesh,
  cutoff = 0.07,
  max.edge = c(0.2, 0.6)
)
plot(mesh_2d, main = "")
```

We now proceed as previously by defining a temporal region and the model
```{r}
t <- seq(from = 0, to = 10, length.out = 11)
kappa <- 10
sigma <- 10
gamma <- 1/200
rho_2d <- 0.3*c(1,1)
alpha <- 1
beta <- 1

op <- spacetime.operators(mesh_space = mesh_2d, time_loc = t,
                          kappa = kappa, sigma = sigma, alpha = alpha,
                          beta = beta, rho = rho_2d, gamma = gamma)
op$plot_covariances(s.ind = 100, t.ind = 5, t.shift = c(-2,0,2))
```

The `spacetime.operators` object has a `plot_covariances` method which can be used to vizualise marginal spatial and 
temporal covariances. The function takes as input `t.ind` which is the index of the location in the time discretization to plot the marginal spatial covariance for, and an input `s.ind` which is the index of the location in the space discretization to show the marginal temporal covariance for. It further takes an input `t.shift` which can be used to plot covariances $C(u(\cdot, t_i), u(\cdot, t_j))$, where $t_i$ is `t[t.ind]` and $t_j$ is `t[t.ind + t.shift]`. For example

We can simulate from the model using `simulate()`: 

```{r}
u <- simulate(op)
```
Let us plot the simulation for a few time points

```{r}
proj <- fm_evaluator(mesh_2d, dims = c(100, 100))
U <- matrix(u, nrow = mesh_2d$n, ncol = length(t))
field1 <- fm_evaluate(proj, field = as.vector(U[,2]))
field2 <- fm_evaluate(proj, field = as.vector(U[,3]))
field3 <- fm_evaluate(proj, field = as.vector(U[,4]))

field1.df <- data.frame(x1 = proj$lattice$loc[,1], x2 = proj$lattice$loc[,2], 
                       u = as.vector(field1), type = "u1")
field2.df <- data.frame(x1 = proj$lattice$loc[,1], x2 = proj$lattice$loc[,2], 
                       u = as.vector(field2), type = "u2")
field3.df <- data.frame(x1 = proj$lattice$loc[,1], x2 = proj$lattice$loc[,2], 
                       u = as.vector(field3), type = "u3")
field.df <- rbind(field1.df, field2.df, field3.df)
ggplot(field.df) + aes(x = x1, y = x2, fill = u) + facet_wrap(~type) + geom_raster() + xlim(0,1) + ylim(0,1) + scale_fill_viridis()


```

We now generate the observations as $Y_i = u(s_i) + \varepsilon_i$, where $\varepsilon_i \sim N(0,\sigma_e^2)$ is Gaussian measurement noise. 
```{r}
n.obs <- 500
obs.loc <- data.frame(x = runif(n.obs),y = runif(n.obs), t = max(t)*runif(n.obs))
A <- op$make_A(cbind(obs.loc$x,obs.loc$y), obs.loc$t)
sigma.e <- 0.01
Y <- as.vector(A%*%u + sigma.e*rnorm(n.obs))
```

To estimate the model parameters based on this data, we can use the `rspde_lme` function. For this, we collect the data in a data frame, that also contanis the spatial locations, and we fit the model:

```{r}
df_2d <- data.frame(Y = as.matrix(Y), x = obs.loc$x, y = obs.loc$y, t = obs.loc$t)
res <- rspde_lme(Y ~ 1, loc = c("x", "y"), loc_time = "t", data = df_2d, model = op, parallel = TRUE)
```

Let us see a summary of the fitted model:

```{r}
summary(res)
```


Let us compare the estimated results with the true values:
```{r}
results <- data.frame(kappa = c(kappa, res$coeff$random_effects[1]), 
                      sigma = c(sigma, res$coeff$random_effects[2]),
                      gamma = c(gamma, res$coeff$random_effects[3]), 
                      rho_1 = c(rho_2d[1], res$coeff$random_effects[4]),
                      rho_2 = c(rho_2d[2], res$coeff$random_effects[5]),
                      sigma.e = c(sigma.e, res$coeff$measurement_error),
                      intercept = c(0, res$coeff$fixed_effects),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

Let us now use our `inlabru` implementation. We first define the model. To this end we need to create a mesh for the time object:

```{r}
mesh_time <- fmesher::fm_mesh_1d(t)
```

```{r}
st_bru_field <- rspde.spacetime(mesh_space = mesh_2d,
        mesh_time = mesh_time, alpha = alpha, beta = beta)
```

Let us now prepare the `data.frame` object:

```{r}
df_bru <- data.frame(y = Y, coord_x = obs.loc$x, coord_y = obs.loc$y, time = obs.loc$t)
```

Now, the component:
```{r}
cmp <- y ~ -1 + Intercept(1) + field(list(space=cbind(coord_x, coord_y), 
                time = time), model = st_bru_field)
```

We are now in a position to fit the model:

```{r}
bru_fit_field <- bru(cmp, data = df_bru)
```

Let us now compare the estimated results (the means of the parameters) with the true values:

```{r}
param_st <- transform_parameters_spacetime(bru_fit_field$summary.hyperpar$mean[2:6], 
                                            st_bru_field)

results <- data.frame(kappa = c(kappa,  param_st[[1]]), 
                      sigma = c(sigma,  param_st[[2]]),
                      gamma = c(gamma,  param_st[[3]]), 
                      rho_1 = c(rho_2d[1], param_st[[4]]),
                      rho_2 = c(rho_2d[2], param_st[[5]]),
                      sigma.e = c(sigma.e, sqrt(1/bru_fit_field$summary.hyperpar$mean[1])),
                      intercept = c(0, bru_fit_field$summary.fixed$mean),
                      row.names = c("True", "Estimate"))
                      
print(results)
```


## Fit with `bounded_rho = FALSE`

In cases where the estimated value of \(\rho\) approaches the upper bound (available in `st_bru$bound_rho`), the model can be re-fitted with `bounded_rho = FALSE`. This removes the bounding constraint on \(\rho\), which can lead to a better fit but may introduce numerical instability. Below, we demonstrate this for both the 1D and spatial examples.

---

### 1D Example with `bounded_rho = FALSE`

Let us first start with the `rspde_lme()` example:

```{r}
op_1d_unbounded <- spacetime.operators(space_loc = s, time_loc = t,
                          kappa = kappa, sigma = sigma, alpha = alpha,
                          beta = beta, rho = rho_1d, gamma = gamma,
                          bounded_rho = FALSE)

res_unbounded <- rspde_lme(y ~ 1, loc = "space", loc_time = "time", 
                      data = df_1d, model = op_1d_unbounded, parallel = TRUE)

results <- data.frame(kappa = c(kappa, res_unbounded$coeff$random_effects[1]), 
                      sigma = c(sigma, res_unbounded$coeff$random_effects[2]),
                      gamma = c(gamma, res_unbounded$coeff$random_effects[3]), 
                      rho = c(rho_1d, res_unbounded$coeff$random_effects[4]),
                      sigma.e = c(sigma.e, res_unbounded$coeff$measurement_error),
                      intercept = c(0, res_unbounded$coeff$fixed_effects),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

Now, the `inlabru` implementation:
```{r}
### Fit with bounded_rho = FALSE
st_bru_unbounded <- rspde.spacetime(space_loc = s, time_loc = t, alpha = 1, 
            beta = 1, bounded_rho = FALSE)

### Fitting the Model
cmp_unbounded <- y ~ -1 + Intercept(1) + field(list(space = space, time = time), 
                                          model = st_bru_unbounded)

bru_fit_unbounded <- bru(cmp_unbounded, data = df_1d)

### Extract and Compare Results
param_unbounded <- transform_parameters_spacetime(
                bru_fit_unbounded$summary.hyperpar$mean[2:5], 
                st_bru_unbounded)

results_unbounded <- data.frame(
  kappa = c(kappa, param_unbounded$kappa),
  sigma = c(sigma, param_unbounded$sigma),
  gamma = c(gamma, param_unbounded$gamma),
  rho = c(rho_1d, param_unbounded$rho),
  sigma.e = c(sigma.e, sqrt(1 / bru_fit_unbounded$summary.hyperpar$mean[1])),
  intercept = c(0, bru_fit_unbounded$summary.fixed$mean),
  row.names = c("True", "Estimate")
)

print(results_unbounded)
```

## References

