---
title: "rSPDE Package"
author: "David Bolin and Alexandre B. Simas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rSPDE Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: bolin19
  title: "The rational SPDE approach for Gaussian random fields with general smoothness"
  author:
  - family: Bolin
    given: David
  - family: Kirchner
    given: Kristin
  container-title: Journal of Computational and Graphical Statistics
  volume: 29
  issue: 2
  page: 274-285
  type: article-journal
  issued:
    year: 2020
- id: lindgren11
  title: "An explicit link between Gaussian fields and Gaussian Markov random fields: the stochastic partial differential equation approach"
  author:
  - family: Lindgren
    given: Finn
  - family: Rue
    given: Håvard
  - family: Lindström
    given: Johan
  container-title: Journal of the Royal Statistical Society. Series B. Statistical Methodology
  volume: 73
  issue: 4
  page: 423--498
  type: article-journal
  issued:
    year: 2011
- id: bolin13
  title: "Spatial Matérn Fields Driven by Non-Gaussian Noise"
  author:
    - family: Bolin
      given: David
  container-title: Scandinavian Journal of Statistics
  volume: 41
  issue: 3
  page: 557-579
  type: article-journal
  issued:
    year: 2013
- id: wallin15
  title: "Geostatistical Modelling Using Non-Gaussian Matérn Fields"
  author:
    - family: Wallin
      given: Jonas
    - family: Bolin
      given: David
  container-title: Scandinavian Journal of Statistics
  volume: 42
  issue: 3
  page: 872-890
  type: article-journal
  issued:
    year: 2015
- id: asar20
  title: "Linear mixed effects models for non‐Gaussian repeated measurement data"
  author:
    - family: Asar
      given: Özgür
    - family: Bolin
      given: David
    - family: Diggle
      given: Peter
    - family: Wallin
      given: Jonas
  container-title: Journal of the Royal Statistical Society. Series C. Applied Statistics
  volume: 69
  issue: 5
  page: 1015-1065
  type: article-journal
  issued:
    year: 2020 
- id: xiong22
  title: "Covariance-based rational approximations of fractional SPDEs for computationally efficient Bayesian inference"
  author: 
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Xiong
    given: Zhen    
  container-title: Journal of Computational and Graphical Statistics
  type: article-journal
  issued:
    year: 2023   
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

set.seed(1)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

# Package Home Page

For more details on the package, along with a brief introduction, function references and several vignettes, go to the [`rSPDE` home page](https://davidbolin.github.io/rSPDE/).

For the source codes, reporting issues, etc., please go the the [`rSPDE` GitHub page](https://github.com/davidbolin/rSPDE).

# Vignettes

We have the following vignettes for the `rSPDE` package:

* [An introduction to the rSPDE package](https://davidbolin.github.io/rSPDE/articles/rSPDE.html)

* [R-INLA implementation of the rational SPDE approach](https://davidbolin.github.io/rSPDE/articles/rspde_inla.html)

* [inlabru implementation of the rational SPDE approach](https://davidbolin.github.io/rSPDE/articles/rspde_inlabru.html)

* [Rational approximation with the rSPDE package](https://davidbolin.github.io/rSPDE/articles/rspde_cov.html)

* [Operator-based rational approximation](https://davidbolin.github.io/rSPDE/articles/rspde_base.html)

* [Building the rSPDE package from source on Mac and Linux](https://davidbolin.github.io/rSPDE/articles/build_source.html)

# Function Manuals

Detailed references for functions including several examples can be found in the [Function References](https://davidbolin.github.io/rSPDE/reference/index.html).
